---
title: "| RESEARCH PROTOCOL\n| \n| First-episode psychosis pharmacotherapy: an antipsychotics comparative effectiveness study \n"
fontsize: 12pt
geometry: margin=1in
output:
  bookdown::html_document2:
    df_print: paged
    toc: yes
    toc_depth: 2
    toc_float: yes
    number_sections: yes
    number_tables: yes
    css: style.css
  word_document:
    reference_docx: ohdsi-protocol-style.docx
  bookdown::pdf_document2:
    keep_tex: yes
    latex_engine: xelatex
    md_extensions: +raw_attribute
    number_sections: yes
    includes:
      before_body: title.tex
header-includes:
- \usepackage[numbers,sort&compress]{natbib}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{caption}
- \usepackage{rotating}
- \usepackage{multirow}
- \usepackage{mwe,tikz}
- \usepackage[percent]{overpic}
- \usepackage{enumitem}
- \usepackage{hyperref}
- \newcolumntype{P}[1]{>{\raggedright\arraybackslash}p{#1}}
- \newcommand{\footerDate}{`r params$date`}
- \input{header.tex}
longtable: yes
mainfont: Arial
bibliography: Protocol.bib
params:
  date: '14-Apr-2025'
  version: 0.0.1
subtitle: 'Version: `r params$version`'
link-citations: yes
csl: bmj.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message = FALSE)
options(kableExtra.latex.load_packages = FALSE)
library(kableExtra)
library(dplyr)
options(knitr.kable.NA = "")
options(knitr.table.format = function() {
  if (knitr::is_latex_output()) {
    "latex"
  } else if (knitr::is_html_output()) {
    "html"
  } else {
    "pipe"
  }
})

latex_table_font_size <- 8
```

# List of Abbreviations

```{r abbreviations, echo=FALSE}
abbreviations <- readr::read_delim(col_names = FALSE, delim = ";", trim_ws = TRUE, file = "
  OHDSI; Observational Health Data Sciences and Informatics
  CDM; Common data model
  ICD-10; International Classification of Diseases, Tenth Revision
  ICD10-CM; International Classification of Diseases, Tenth Revision, Clinical Modification
  DALY; Disability-Adjusted Life Years
  FEP; First-Episode Psychosis
  SGA; Second-generation antipsychotic
  FGA; First-generation antipsychotic
  AP; Antipsychotic
")

tab <- kable(abbreviations, col.names = NULL, linesep = "", booktabs = TRUE)

if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = c("striped", "hold_position"), 
                        font_size = latex_table_font_size)
} else {
  tab %>% kable_styling(bootstrap_options = "striped")
}
```

\clearpage

# Responsible Parties

## Investigators

## Disclosures

This study is undertaken within Observational Health Data Sciences and Informatics (OHDSI), an open collaboration.
\clearpage

# Abstract

**Background and Significance**: 
Schizophrenia is a severe neuropsychiatric disorder which often lead to a significant functional and cognitive impairment. Although its prevalence is relatively low [1], schizophrenia is one of the biggest causes of DALYs worldwide [2]. The term 'first-episode schizophrenia' is often used interchangeably with the term 'first-episode psychosis', since distinguishing former from the latter is a complex and often impossible clinical task, because clinical picture and further manifestations of the disorder may be multimodal and evolving. Studies show that interventions at the earliest stages of the disorder are important for  ameliorating the course of illness [3, 4]. Treatment with antipsychotics is one of such interventions [5]. However, the choice of medication for treatment of FEP requires further investigation. Current clinical practice guidelines mention the heterogeneity of clinical trial designs and limited number of head-to-head comparisons as factors preventing from creating an evidence-based ranking of SGAs and FGAs [6].

**Study Aims**: 
This study aims at creating a head-to-head comparison framework for the effectiveness several commonly used FGAs and SGAs when used in treating patients with first-episode psychoses.

**Study Description**: 
We will conduct a large-scale, systematic, observational study for each of the pairwise comparisons of 9 drugs of interest (Risperidone, Haloperidol, Aripiprazole, Brexpiprazole, Olanzapine, Quetiapine, Ziprazidone, Lurazidone, Paliperidone). This list includes the most common antipsychotic agents used for treatment of the first psychotic episode as per the results of OHDSI "Phenotype Phebruary 2025" initiative [7].

* **Population**: 
Patients 18-40 years old experiencing a first-in-a-lifetime of psychotic episode, and with no record of treatment non-adherence during the diagnosis period who are first-time users of a specific AP. For a diagnosis setting, we will require either a recorded inpatient visit or 2 subsequent visits within 30 days. We will exclude patients who had a diagnosis of an organic or drug-induced mental disorder to control for diagnosis recording differences across datasets.  

* **Comparators**:
We defined drugs by a respective ingredient or, in the case of Quetiapine, of a dose of 100 mg or more. First exposure  to one of the drugs on the list will be an indexing event, and we will not allow for polypharmacotherapy with APs by censoring on exposure to any other AP unless it happens between the 2 visits requires in the "Population" section.

* **Outcomes**:
We will use rehospitalization with a diagnosis of a psychotic disorder as a proxy for drug effectiveness. As mentioned above, relapse prevention is one of the key targets in treatment of first-episode psychosis. Rehospitalization as an outcome encompasses both inherent drug effectiveness as well as such factors as side effects influencing treatment adherence.

* **Design**: 
Observational, active-comparator, new-user cohort study.

* **Timeframe**: 
We will consider Time-at-Risk window to start from the 31st day from cohort start (initial drug exposure) and last for 180 days.

\clearpage

# Amendments and Updates



# Milestones



# Rationale and Background
In the research field of psychiatry, measuring treatment effectiveness and outcomes and providing clinicians a clear roadmap for treatment are challenging tasks. This is, among other things, related to low perceived severity of the disorders, perceived treatment ineffectiveness and stigma among patients [8]. Such complications to the research process result in difficulties when stratifying treatment options in the clinical guidelines, giving the practitioners guidance on the general principles, but leading to inconsistent treatment and care patterns across the globe [9-11].
# Study Objectives

# Research Methods

## Study Design

This study uses `CohortMethod` [@Schuemie2020-fa].

## Data Sources

## Study Population

## Exposure Comparators

## Outcomes {#outcomes}

## Analysis

# Sample Size and Study Power {#sample-size}

# Strengths and Limitations {#strengths-limitations}

# Protection of Human Subjects

# Management and Reporting of Adverse Events and Adverse Reactions

# Plans for Disseminating and Communicating Study Results

\clearpage

# References {-}
```{r references, echo=FALSE}
references <- readr::read_delim(col_names = FALSE, delim = "|", trim_ws = TRUE, file = "
  1| Moreno-Küstner B, Martín C, Pastor L. Prevalence of psychotic disorders and its association with methodological issues. A systematic review and meta-analyses. PLoS One. 2018;13(4):e0195687. Published 2018 Apr 12. doi:10.1371/journal.pone.0195687
  2| GBD 2019 Mental Disorders Collaborators. Global, regional, and national burden of 12 mental disorders in 204 countries and territories, 1990-2019: a systematic analysis for the Global Burden of Disease Study 2019. Lancet Psychiatry. 2022;9(2):137-150. doi:10.1016/S2215-0366(21)00395-3
  3| Nordentoft M, Rasmussen JO, Melau M, Hjorthøj CR, Thorup AA. How successful are first episode programs? A review of the evidence for specialized assertive early intervention. Curr Opin Psychiatry. 2014;27(3):167-172. doi:10.1097/YCO.0000000000000052
  4| Correll CU, Galling B, Pawar A, et al. Comparison of Early Intervention Services vs Treatment as Usual for Early-Phase Psychosis: A Systematic Review, Meta-analysis, and Meta-regression. JAMA Psychiatry. 2018;75(6):555-565. doi:10.1001/jamapsychiatry.2018.0623
  5| Alvarez-Jiménez M, Parker AG, Hetrick SE, McGorry PD, Gleeson JF. Preventing the second episode: a systematic review and meta-analysis of psychosocial and pharmacological trials in first-episode psychosis. Schizophr Bull. 2011;37(3):619-630. doi:10.1093/schbul/sbp129
  6| Keepers GA, Fochtmann LJ, Anzia JM, et al. The American Psychiatric Association Practice Guideline for the Treatment of Patients With Schizophrenia. Am J Psychiatry. 2020;177(9):868-872. doi:10.1176/appi.ajp.2020.177901
  7| https://results.ohdsi.org/app/26_PhenotypePhebruary2025
  8| Viana MC, Kazdin AE, Harris MG, et al. Barriers to 12-month treatment of common anxiety, mood, and substance use disorders in the World Mental Health (WMH) surveys. Int J Ment Health Syst. 2025;19(1):6. Published 2025 Feb 9. doi:10.1186/s13033-024-00658-2
  9| Vigo DV, Stein DJ, Harris MG, et al. Effective Treatment for Mental and Substance Use Disorders in 21 Countries. JAMA Psychiatry. 2025;82(4):347-357. doi:10.1001/jamapsychiatry.2024.4378
  10| Martin A, Bessonova L, Hughes R, et al. Systematic Review of Real-World Treatment Patterns of Oral Antipsychotics and Associated Economic Burden in Patients with Schizophrenia in the United States. Adv Ther. 2022;39(9):3933-3956. doi:10.1007/s12325-022-02232-z
  11| Kohn R, Saxena S, Levav I, Saraceno B. The treatment gap in mental health care. Bull World Health Organ. 2004;82(11):858-866.
")

tab <- kable(references, col.names = NULL, linesep = "", booktabs = TRUE)

if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = c("striped", "hold_position"), 
                        font_size = latex_table_font_size)
} else {
  tab %>% kable_styling(bootstrap_options = "striped")
}
```

<div id="refs"></div>

\clearpage

\centerline{\Huge Appendix}

# (APPENDIX) Appendix {-}

# Exposure Cohort Definitions


# Outcome Cohort Definitions

# Negative Control Concepts {#negative-controls}